Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-01-22 paper/paper.tex
put cases in appendix

\section{Type-theoretic Characterizations of Language Classes}
Classically, formal language theory is closely related to automata theory. In the styles of Chomsky and Sch\"utzenberger, there is a hierarchy of language classes and the types of automata that recognize them. Inside of the type theory of \cref{sec:typetheory}, we can characterize these language classes; moreover, we demonstrate equivalence of each class to the associated automaton as an isomorphism within our logic.

Note, predicate BI with existential quantifiers is already enough to define
formulae that correspond to arbitrary recursively enumerable languages
since we can simulate the tape of a Turing machine using two
stacks. I.e., we can define a set $S$ of lists of tape symbols, and
then define a formula that first ``copies'' the input onto one stack, then executes

@@

We reflect on this expressive power to motivate why we put heavy syntactic restrictions on our presentation of certain types of automata. For instance, when defining a pushdown automata, we ought to be very careful about how the set representing the stack can be accessed and added to. Careless definitions may accidentally endow us with too much expressive power, launching us further up in the Chomsky hierarchy. To begin, let us focus only on language classes, and their associated automata whose parses are definable without any mention to the inuitionistic context. In the context of automata, this refers to the class of automata definable without access to any memory --- i.e.\ finite state machines.

\subsection{Finite State Machines}
We may encode a finite state machine via a particular syntactic form for grammars. A \emph{nondeterministic finite automaton} (NFA) is defined as a mutual \footnote{For convenience we refer to a mutual fixed point here, even though we only take single fixed points as primitive. As in Beki\'c's theorem, we may decompose our mutual fixed points into a series of single fixed points.} fixed point in the following manner,

\steven{Is there a better way to present this?}
\[
  NFA ::= \mu \begin{pmatrix}
        Y_{s} = \langle Q \rangle \\
        Y_{1} = \langle Q \rangle \\
        \vdots \\
        Y_{e} = I
        \end{pmatrix}~\textbf{in}~Y_{s}
\]
where
\[
  Q ::= c \otimes Y ~|~ Y ~|~ Q \oplus Q
\]
and $Y \in \{ Y_{s}, Y_{1}, \dots, Y_{N} \}$. Intuitively, we read each $Y_{j}$ as the label for a state, and the definitions in form $Q$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$ an $\epsilon$-transition. Without loss of generality, $Y_{s}$ is reserved for a single distinguished start state, and $Y_{e}$ is reserved for a single distinguished accepting state. A disjunction\footnote{There is an admissible notion of coproduct making use of the primitive dependent pairing --- i.e. $A \oplus B := \LinPiTy b \Bool {\textbf{if}~b~A~B}$. We will utilize this without much comment to represent the usual notion of alternation of grammars.} of transitions in an NFA grammar corresponds to multiple outgoing edges of a given state in an automaton.

Let $\hat{Y}_{k}$ refer to the syntactic term on the right hand side of the definition of $Y_{k}$ inside of an NFA $N$. We say that $A$ is a \emph{deterministic finite automaton} (DFA) if there are no $\epsilon$-transitions and for every definition $\hat{Y}_{k}$ the $c_{j}$ are distinct guards for each produced state. \steven{Is this the right notion of determinism? Perhaps, it is is better to define determinism as something like: there is at most one derivation for all $j, k$, for all $a \in \Sigma$ we have $x : Y_{k}, y : a \vdash z : Y_{j}$.}

With this setting for finite automata, we can now internalize some classical theorems inside of our formal system. First, let us describe a version of Thompson's construction \cite{??} where we show that any regular expression can be recognized by some NFA\@. Moreover, we will show that this NFA is strongly equivalent to the original grammar.

\subsection{Equivalence Between Regular Expressions and Finite Automata}

\steven{Define regular expressions}
\steven{Move proofs to appendix, leave the exposition and the theorem statements.}

\subsubsection{Grammar to NFA}

We will define the recognizing NFA of a regular expression inductively. To handle the atomic grammars --- literals and the empty grammar --- we define the following,
\[
  NFA(c) =
  \mu \begin{pmatrix}
        Y_{s} = c \otimes Y_{e} \\
        Y_{e} = I
  \end{pmatrix}~\textbf{in}~Y_{s}
\]
\[
  NFA(I) =
  \mu \begin{pmatrix}
        Y_{s} = Y_{e} \\
        Y_{e} = I
  \end{pmatrix}~\textbf{in}~Y_{s}
\]

@@

We can now inductively define some NFA combinators and show that they preserve grammar equivalence. First begin with the disjunction of two NFAs. hat is, assuming that $g \cong NFA(g)$ and $g' \cong NFA(g')$, we define an NFA that recognizes $g \oplus g'$. Let the states in $NFA(g)$ be represented by $X$'s and the states in $NFA(g')$ be represented by $Z$'s,

\[
  NFA(g \oplus g') =
  \mu \begin{pmatrix}
        Y_{s} =  X_{s}' \oplus Z_{s}' \\
        X_{s}', X_{1}', \dots = \dots \\
        X'_{e} = Y_{e} \\
        Z_{s}', Z_{1}, \dots = \dots \\
        Z_{e}' = Y_{e} \\
        Y_{e} = I
  \end{pmatrix}~\textbf{in}~Y_{s}
\]

For each defintion $X_{j}$ in $NFA(g)$ we create an analogous definition $X_{j}'$ inside of $NFA(g \oplus g')$ with the syntactic equality $X_{j}'$ of $X_{j}$ for $j \neq e$, except we replaec $X_{k}$ with $X_{k}'$ in each production. Lastly, $X_{e}'$ has an $\epsilon$-transition to the accepting state $Y_{e}$. Likewise, we build analogous defintions for the states $Z_{j}$ in $NFA(g')$, then we flatten all of these defintions into one mutally recursive list.

To show the strong equivalence of $NFA(g \oplus g')$ and $g \oplus g'$, we will demonstrate an isomorphism between them with two proof terms in our logic that mutually invert each other. To define a map $NFA(g \oplus g') \to g \oplus g'$ we will use a multi-fold over all the mutually recursive defintions of $NFA(g \oplus g')$. Even though multiple folding is not taken as primitive, using Beki\'c's theorem we can define a family of multifold terms to serve as elimination forms for our admissible multiple fixed points. That is, rules of the form,

\[
  \inferrule{\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n})~\textbf{in}~X_{k} \\
             \Gamma ; x_{1} : \simulsubst {A_{1}}{B_{1}/X_{1},\dots,B_{n}/X_{n}} \vdash e_{1} : B_{1} \\
             \dots \\
             \Gamma ; x_{n} : \simulsubst {A_{n}}{B_{1}/X_{1},\dots,B_{n}/X_{n}} \vdash e_{n} : B_{n}
  }{\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
\]

Let the substitution $\{B_{1}/X_{1},\dots,B_{n}/X_{n} \}$ be given by $\gamma$. Additionally, let the substitution $\{X_{s}'/X_{s}, X_{e}'/X_{e}, X_{1}'/X_{1}, \dots \}$ be given by $\delta_{X}$. Likewise define $\delta_{Z}$. We can then characterize the map from $NFA(g \oplus g')$ with the following term,

@@

For each of the variables above, we simplify the type and provide the actual proof term. For instance, $\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$, so the derivation for $p_{s}$ is just the identity. We do this for all types and terms.

Let $\hat{X}_{k}$ refer to the syntactic term on the right hand side of the definition of $X_{k}$ inside of $NFA(g)$. Likewise define $\hat{Z}_{k}$. Thus, syntactically $X_{k}' = \simulsubst {\hat{X}_{k}} {\delta_{X}}$ for $k \neq e$.

\begin{itemize}
  \item $y_{s}$ --- $\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$ and $p_{s} = y_{s} : g \oplus g'$
  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = X_{s}$ and $p_{x,s} = \phi^{-1}(x_{s}') : g$
  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = X_{j}$ and $p_{x, j} = x_{j}'$
  \item $x_{e}'$ --- $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{x,e} = x_{e}' : X_{e} = I$
  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = Z_{s}$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = Z_{j}$ and $p_{z, j} = z_{j}'$
  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = z_{e}' : Z_{e} = I$
  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
\end{itemize}

where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively. The map in the other direction is give as,

@@

To see this consider a case distinction on $e : NFA(g \oplus g')$. \steven{I could type out the equality proof term, but it blows up in size and doesn't seem like a useful way to capture this iso. I have a commented out version of it above. Going to refactor tomorrow to something more sane, as these as are the crux of the proofs of equality} \steven{In any case, the NFA-to-grammar direction of these isos should all follow from some $\eta$-rule on the structure of the type for the start state, right?}
\pedro{What I like doing in these cases is either presenting the high-level structure of the proof, e.g., ``by applying the rules <...> we can conclude'', or present the full proof in the appendix.
  Since in this case it feels like it won't be a particularly illuminating proof, I suggest doing
the former.}
Without loss of generality assume that $e = \textbf{in}_{X_{s}'} u$. Then, $\mathsf{mfold}(e) = \textbf{in}_{g}(\phi^{-1}u)$ and passing this through the above case statement results in $\textbf{in}_{{X_{s}'}}(\phi(\phi^{{-1}}u)) = \textbf{in}_{{X_{s}'}}(u) = e$. Further,

@@

for very similar reasons. \steven{How do you actually unpack what this fold term does?}

We now construct a combinator for the sequencing of two NFAs.

\pedro{I feel like that for the sake of organizing the paper, since the proof is non-trivial,
there should be a theorem environment with an accompanying proof environment}

@@

Thus we have defined a second model $\gluedNL,\gluedL$ of the type
theory, with a morphism to the syntactic model. By the initiality of
the syntactic model, there is a morphism back to the glued model which
is a section of this projection. For purely positive types, we can
observe by induction on the type structure that the total space of the
family is isomorphic to the set-theoretic semantics. Further, the
projection above defines a section of the families. Finally, by
induction on formulae again we see that this projection is also a
section and therefore an isomorphism, establishing the canonicity
result as stated in Section\ref{section:canonicity}.
\end{document}

>>>>

\section{Type-theoretic Characterizations of Language Classes}
Classically, formal language theory is closely related to automata theory. In the styles of Chomsky and Sch\"utzenberger, there is a hierarchy of language classes and the types of automata that recognize them. The type theory of \cref{sec:typetheory} serves as a syntax to reason about formal grammars; moreover, we describe notions of automata into this formal system via syntactic forms that describe an automaton as a grammar. This method allows us to internalize many classic automata-theoretic results as proof terms in our type theory.

Note, a predicate logic over LNL with existential quantifiers is already enough to define
formulae that correspond to arbitrary recursively enumerable languages ---
since we can simulate the tape of a Turing machine using two
stacks. I.e., we can define a set $S$ of lists of tape symbols, and
then define a formula that first ``copies'' the input onto one stack, then executes

@@

We reflect on this expressive power to motivate why we put heavy syntactic restrictions on our presentation of certain classes of automata. For instance, when defining a pushdown automata, we could use the intuitionistic context to represent the stack, but we ought to be very careful about how this set is accessed and added to. Careless definitions may accidentally endow us with too much expressive power, launching us further up in the Chomsky hierarchy. To begin, let us focus only on language classes, and their associated automata, whose parses are definable without any mention to the inuitionistic context. When viewed as a machine, this refers to the class of automata definable without access to any memory --- i.e.\ finite state machines.

\subsection{Finite State Machines}
A finite state machine may be expressed via a particular syntactic form for grammars. A \emph{nondeterministic finite automaton} (NFA) is defined as a mutual\footnote{For convenience we refer to a mutual fixed point here, even though we only take single fixed points as primitive. As in Beki\'c's theorem, we may decompose our mutual fixed points into a series of single fixed points.} fixed point in the following manner,

\[
  NFA ::= \mu \begin{pmatrix}
        Y_{s} = \langle Q \rangle \\
        Y_{1} = \langle Q \rangle \\
        \vdots \\
        Y_{e} = I
        \end{pmatrix}~\textbf{in}~Y_{s}
\]
where
\[
  Q ::= c \otimes Y ~|~ Y ~|~ Q \oplus Q
\]
and $Y \in \{ Y_{s}, Y_{1}, \dots, Y_{N} \}$. Intuitively, we read each $Y_{j}$ as the label for a state, and the definitions in form $Q$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$ denotes an $\epsilon$-transition. Without loss of generality, $Y_{s}$ is reserved for a single distinguished start state, and $Y_{e}$ is reserved for a single distinguished accepting state. A disjunction\footnote{There is an admissible notion of coproduct making use of the primitive dependent pairing --- i.e. $A \oplus B := \LinPiTy b \Bool {\textbf{if}~b~A~B}$. We will utilize this without much comment to represent the usual notion of alternation of grammars.} of transitions in an NFA grammar corresponds to multiple outgoing edges of a given state in an automaton.

Let $\hat{Y}_{k}$ refer to the syntactic term on the right hand side of the definition of $Y_{k}$ inside of an NFA $N$. We say that $N$ is a \emph{deterministic finite automaton} (DFA) if there are no $\epsilon$-transitions and for every definition $\hat{Y}_{k}$ the $c_{j}$ are distinct guards for each produced state.
% \steven{Is this the right notion of determinism? Perhaps, it is is better to define determinism as something like: there is at most one derivation for all $j, k$, for all $a \in \Sigma$ we have $x : Y_{k}, y : a \vdash z : Y_{j}$.}

With this setting for finite automata, we can now internalize some classical theorems inside of our formal system. First, let us describe a version of Thompson's construction \cite{?} where we show that any regular expression can be recognized by some NFA\@. Moreover, we will show that this NFA is strongly equivalent to the original grammar.

\subsection{Equivalence Between Regular Grammars and Finite Automata}

The class of \emph{regular} grammars is given as follows:

\[
Reg ::= 0 ~|~ I ~|~ c ~|~ Reg \oplus Reg ~|~ Reg \otimes Reg ~|~ Reg^{*}
\]
where $c \in \mathcal{A}$ is a character literal drawn from a fixed alphabet, and the Kleene star $\cdot^{*}$ is derived as an admissible operation using the fixed point operator --- $g^{*} = \mu X.I \oplus (g \otimes X)$. \steven{This definition is maybe fair to assume. Also maybe better suited for a section earlier in the paper}

\begin{theorem}
  \label{thm:thompson}
  For $g$ a regular grammar, there is an NFA $N$ that recognizes the same language as $g$. Further, $N$ and $g$ are isomorphic.
\end{theorem}

\begin{proof}
  We construct the recognizing NFA $N$ inductively. To handle the atomic grammars --- literals and the empty grammar --- we define the following,

\[
  NFA(c) =
  \mu \begin{pmatrix}
        Y_{s} = c \otimes Y_{e} \\
        Y_{e} = I
  \end{pmatrix}~\textbf{in}~Y_{s}
\]
\[
  NFA(I) =
  \mu \begin{pmatrix}
        Y_{s} = Y_{e} \\
        Y_{e} = I
  \end{pmatrix}~\textbf{in}~Y_{s}
\]

@@

We can now inductively define some NFA combinators and show that they preserve grammar equivalence. First begin with the disjunction of two NFAs. That is, assuming that $g \cong NFA(g)$ and $g' \cong NFA(g')$, we define an NFA that recognizes $g \oplus g'$. Let the states in $NFA(g)$ be represented by $X$'s and the states in $NFA(g')$ be represented by $Z$'s. Further let $\hat{X}_{k}$ refer to the syntactic term on the right hand side of the definition of $X_{k}$ inside of $NFA(g)$. Likewise define each $\hat{Z}_{k}$.

\[
  NFA(g \oplus g') =
  \mu \begin{pmatrix}
        Y_{s} =  X_{s}' \oplus Z_{s}' \\
        X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
        X'_{e} = Y_{e} \\
        Z_{j}' = \simulsubst {\hat{Z}_{j}} {\delta_{Z}}, j \neq e \\
        Z_{e}' = Y_{e} \\
        Y_{e} = I
  \end{pmatrix}~\textbf{in}~Y_{s}
\]
where the substitution $\{X_{s}'/X_{s}, X_{e}'/X_{e}, X_{1}'/X_{1}, \dots \}$ is given by $\delta_{X}$. Likewise define $\delta_{Z}$.
In the construction of $NFA(g \oplus g')$ we replicate each defintion $X_{j}$ in $NFA(g)$ with an analogous definition $X_{j}'$ inside of $NFA(g \oplus g')$, except we replace $X_{k}$ with $X_{k}'$ in each production. Lastly, $X_{e}'$ has an $\epsilon$-transition to the accepting state $Y_{e}$. Likewise, we build analogous defintions for the states $Z_{j}$ in $NFA(g')$, then we flatten all of these defintions into one mutally recursive list. This has the effect of disjunctively tying together $NFA(g)$ and $NFA(g')$ into one automaton.

To show the strong equivalence of $NFA(g \oplus g')$ and $g \oplus g'$, we will demonstrate an isomorphism between them with two proof terms in our logic that mutually invert each other. To define a map $NFA(g \oplus g') \to g \oplus g'$ we will use a multi-fold over all the mutually recursive defintions of $NFA(g \oplus g')$. Note, even though multiple folding is not taken as primitive, using Beki\'c's theorem we can define a family of multifold terms to serve as elimination forms for our admissible multiple fixed points. That is, rules of the form,

\[
  \inferrule{\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n})~\textbf{in}~X_{k} \\
             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
             \dots \\
             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
  }{\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
\]

are admissible, where the substitution $\{B_{1}/X_{1},\dots,B_{n}/X_{n} \}$ is given by $\gamma$. In this instance, we wish to simultaneously fold over the mutually recursive definitions comprising $NFA(g \oplus g')$. We may think of this as an eliminator that replaces each state inside of $NFA(g \oplus g')$ with a tail-recursive call that will ultimately return a term of type $g \oplus g'$. The structure of such a derivation is given below,

@@

For each of the variables above, we simplify the type and provide the actual proof term. For instance, $\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$, so the derivation for $p_{s}$ is just the identity. The rest of the terms are given below,
\begin{itemize}
  \item $y_{s}$ --- $\simulsubst {(X_{s}' \oplus Z'_{s})} \gamma = g \oplus g'$ and $p_{s} = y_{s} : g \oplus g'$
  \item $x_{s}'$ --- $\simulsubst {(\simulsubst {\hat{X}_{s}} {\delta_{X}})}{\gamma} = X_{s}$ and $p_{x,s} = \phi^{-1}(x_{s}') : g$
  \item $x_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{X}_{j}} {\delta_{X}})}{\gamma} = X_{j}$ and $p_{x, j} = x_{j}'$
  \item $x_{e}'$ --- $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{x,e} = x_{e}' : X_{e} = I$
  \item $z_{s}'$ --- $\simulsubst {(\simulsubst {\hat{Z}_{s}} {\delta_{Z}})}{\gamma} = Z_{s}$ and $p_{z,s} = \psi^{-1}(z_{s}') : g'$
  \item $z_{j}', j \not \in \{s, e\}$ --- $\simulsubst {(\simulsubst {\hat{Z}_{j}} {\delta_{Z}})}{\gamma} = Z_{j}$ and $p_{z, j} = z_{j}'$
  \item $z_{e}'$ ---  $\simulsubst {Y_{e}}{\gamma} = I$ and $p_{z,e} = z_{e}' : Z_{e} = I$
  \item $y_{e}$ --- $\simulsubst {I} {\gamma} = I$ and $p_{e} = y_{e} : I$
\end{itemize}

where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively. This finshes the construction of a parse transformer $NFA(g \oplus g') \to g \oplus g'$ The map in the other direction is given the derivation,

@@

% \steven{I could type out the equality proof term, but it blows up in size and doesn't seem like a useful way to capture this iso. I have a commented out version of it above. Going to refactor tomorrow to something more sane, as these as are the crux of the proofs of equality} \steven{In any case, the NFA-to-grammar direction of these isos should all follow from some $\eta$-rule on the structure of the type for the start state, right?}
% \pedro{What I like doing in these cases is either presenting the high-level structure of the proof, e.g., ``by applying the rules <...> we can conclude'', or present the full proof in the appendix.
  % Since in this case it feels like it won't be a particularly illuminating proof, I suggest doing
% the former.}
    %

By breaking into a case distinction on $e$ and then applying the $\eta$-rule for coproducts, the above proof will simplify into two cases where $\phi$ will be inverted by $\phi^{{-1}}$ and $\psi$ will be inverted by $\psi^{-1}$, ultimately cancelling to give the identity map. The other direction below simplifies in a similar manner. \steven{Most concise list for this?}

% Without loss of generality assume that $e = \textbf{in}_{X_{s}'} u$. Then, $\mathsf{mfold}(e) = \textbf{in}_{g}(\phi^{-1}u)$ and passing this through the above case statement results in $\textbf{in}_{{X_{s}'}}(\phi(\phi^{{-1}}u)) = \textbf{in}_{{X_{s}'}}(u) = e$. Further,

@@

This concludes the disjunctive combination of two NFAs. Because they follow a very similar style of argumentation, we include the cases for sequencing and Kleene star in \cref{sec:appendixnfa}.

\end{proof}

@@

Thus we have defined a second model $\gluedNL,\gluedL$ of the type
theory, with a morphism to the syntactic model. By the initiality of
the syntactic model, there is a morphism back to the glued model which
is a section of this projection. For purely positive types, we can
observe by induction on the type structure that the total space of the
family is isomorphic to the set-theoretic semantics. Further, the
projection above defines a section of the families. Finally, by
induction on formulae again we see that this projection is also a
section and therefore an isomorphism, establishing the canonicity
result as stated in Section\ref{section:canonicity}.

\section{NFA stuff}
\label{sec:appendixnfa}

We provide the final two cases of \cref{thm:thompson}.

\begin{proof}
We now construct a combinator for the sequencing of two NFAs.

\end{proof}

\end{document}
